---
title: "Exploring Relationships Between Extreme Poverty Rates and Inflation Rates in the World"
author: "Tarj Mecwan, Bernice Lau, Rachel Hartfelder, Haley Tran"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-fold: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## Dependencies

To start, we need to load the `tidyverse` and `knitr` packages into RStudio.

```{r}
#| message: false

library(tidyverse)
library(knitr)
```

## Datasets

Next, we will read in our datasets!

```{r}
#| message: false

poverty <- read_csv("data/poverty.csv")
inflation <- read_csv("data/inflation.csv")
```

Let's examine the `poverty.csv` dataset.

```{r}
head(poverty) |> 
  kable()
```

Here's what the `inflation.csv` dataset looks like.

```{r}
head(inflation) |> 
  kable()
```

**TODO**: Detailed data and variable description. (Rachel)

-   Summarize the poverty dataset (Extreme Poverty Rate, less than $2.15/day)

-   Summarze the inflation dataset (Inflation, annual %)


### Data Cleaning

**TODO**: Discussion of data cleaning process and decisions. (Tarj)

```{r}
# Filter the datasets to include only the overlapping years (1961 to 2023):

inflation_filtered <- inflation |>
  select(country, `1961`:`2023`)

poverty_filtered <- poverty |>
  select(country, `1961`:`2023`)

```

```{r}
# 1.2 Pivoting Longer
# Pivot the year columns to a single column in both datasets

inflation_long <- inflation_filtered |>
  pivot_longer(cols = `1961`:`2023`, 
               names_to = "year", 
               values_to = "inflation_rate") |>
  mutate(year = as.numeric(year))

poverty_long <- poverty_filtered |>
  pivot_longer(cols = `1961`:`2023`, 
               names_to = "year", 
               values_to = "poverty_rate") |>
  mutate(year = as.numeric(year))

```

```{r}
# 1.3 Joining Datasets
# Join the datasets on country and year

poverty_inflation_data <- inflation_long |>
  inner_join(poverty_long, by = c("country", "year"))

```

```{r}
# Handle missing values

poverty_inflation_data <- poverty_inflation_data |>
  drop_na()

```

```{r}
head(poverty_inflation_data) |>
  kable()
```
**TODO**: Detailed data and variable description. (Rachel)

-   Summarize the joined dataset

## Hypothesis

**TODO**: Hypothesized relationship between the variables (and any outside references). (Haley)
